Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-01-19 manuscript/HIAs.qmd
filling in area stats

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliations: 
        - id: some-tech
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - SEEA EA
  - alien species
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
# remove this execute part to actually run the analysis
execute:
  echo: false
  include: false
  eval: false
---

@@

```{r prepData, eval=F}
# Set up conditional file paths
# The data is too big to be stored on GitHub
dir <- substr(getwd(), 1,2)

@@

```{r}
# read cached rds file
naturetypes <- readRDS("../data/surveyData.RDS")
```

```{r getNorwayOutline}
outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp") |>
  st_transform(25832)
```

```{r getMunicipalityOutine}
path <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")

@@

# function to make sure that mutipolygons are returned 
ensure_multipolygons <- function(X) {
    tmp1 <- tempfile(fileext = ".gpkg")
    tmp2 <- tempfile(fileext = ".gpkg")
    st_write(X, tmp1)
    gdalUtilities::ogr2ogr(tmp1, tmp2, f = "GPKG", nlt = "MULTIPOLYGON")
    Y <- st_read(tmp2)
    st_sf(st_drop_geometry(X), geom = st_geometry(Y))
}

@@

# subset of the three target municipalities
muni3 <- muni |>
  filter(kommunenummer %in% c("3020", #Nordre Follo
                              "5001", # Trondheim
                              "5418"  # Målselv
                              )) |>
  mutate(Municipality = case_when(
    kommunenummer == "3020" ~ "Nordre Follo",
    kommunenummer == "5001" ~ "Trondheim",
    kommunenummer == "5418" ~ "Målselv"
  ))
# To crop EDM, I need the three municipalities seprately. 
nf <- muni3 |>
  filter(kommunenummer == "3020")
tr <- muni3 |>
  filter(kommunenummer == "5001")
me <- muni3 |>
  filter(kommunenummer == "5418")

# Make map to show where the three municipalities are
# First, get centroids
centroids <- muni3 |>
  st_centroid()

@@

# Simplified coastline
outline2 <- outline |>
  st_intersection(muni3)

@@

# Plot (could perhaps add the EDM here as well)
muniPlot <- tm_shape(muni3)+
    tm_borders()+
  tm_facets(by = "Municipality",
            ncol=2)+
  tm_shape(outline2)+
    tm_fill(col = "lightgreen",
             alpha=.4)+
  tm_shape(ocean)+
    tm_polygons(col = "skyblue",
                border.col = "black")+
  tm_shape(dk)+
    tm_polygons(col = "grey",
                alpha=.8)+
  tm_shape(nature3)+
  tm_polygons(col = "red",
              border.col = "red",
              lwd=3)

@@

```{r getMireData}
file <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
# This is a big TIF, so I some lazy evalyation on the starts proxy 
nf_mire  <- stars::read_stars(file, proxy=T) |>
  st_crop(nf) |>
  st_as_stars()

@@

# This is not working as it puts the cropped raster into the sea
#tmap_mode("view")
#tm_shape(st_downsample(nf_mire , 10))+
#  tm_raster(alpha=.8
#            )+
#  tm_shape(outline)+
#  tm_borders()

@@

@@

>>>>

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliations: 
        - id: some-tech
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - SEEA EA
  - alien species
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
# remove this execute part to actually run the analysis
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

@@

```{r prepData}
#| eval: false

# Set up conditional file paths
# The data is too big to be stored on GitHub
dir <- substr(getwd(), 1,2)

@@

```{r readSurvey}
#| eval: true

# read cached rds file
naturetypes <- readRDS("../data/surveyData.RDS")
```

```{r getNorwayOutline}
#| eval: true
outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp") |>
  st_transform(25832)
```

```{r prepPolygons}
#| eval: true

path <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")

@@

# function to make sure that multipolygons are returned 
ensure_multipolygons <- function(X) {
    tmp1 <- tempfile(fileext = ".gpkg")
    tmp2 <- tempfile(fileext = ".gpkg")
    st_write(X, tmp1)
    gdalUtilities::ogr2ogr(tmp1, tmp2, f = "GPKG", nlt = "MULTIPOLYGON")
    Y <- st_read(tmp2)
    st_sf(st_drop_geometry(X), geom = st_geometry(Y))
}

@@

# subset of the three target municipalities
muni3 <- muni |>
  filter(kommunenummer %in% c("3020", #Nordre Follo
                              "5001", # Trondheim
                              "5418"  # Målselv
                              )) |>
  mutate(Municipality = case_when(
    kommunenummer == "3020" ~ "Nordre Follo",
    kommunenummer == "5001" ~ "Trondheim",
    kommunenummer == "5418" ~ "Målselv"
  ))

# To crop EDM, I need the three municipalities seprately. 
nf <- muni3 |>
  filter(kommunenummer == "3020")
tr <- muni3 |>
  filter(kommunenummer == "5001")
me <- muni3 |>
  filter(kommunenummer == "5418")

@@

# Simplified coastline / terrestrial area
terrestrial <- outline |>
  st_intersection(muni3)

@@

# claculate stats

# terrestrial area
terrestrial <- terrestrial |>
  mutate(area_t = geometry |> st_area())

```{r makeMap}
# Make map to show where the three municipalities are
# First, get centroids
centroids <- muni3 |>
  st_centroid()

# Plot (could perhaps add the EDM here as well)
muniPlot <- tm_shape(muni3)+
    tm_borders()+
  tm_facets(by = "Municipality",
            ncol=2)+
  tm_shape(terrestrial)+
    tm_fill(col = "lightgreen",
             alpha=.4)+
  tm_shape(ocean)+
    tm_polygons(col = "skyblue",
                border.col = "black")+
  tm_shape(dk)+
    tm_polygons(col = "grey",
                alpha=.8)+
  tm_shape(nature3)+
  tm_polygons(col = "red",
              border.col = "red",
              lwd=3)

@@

```{r getMireData}
file <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
# This is a big TIF, so I some lazy evalyation on the starts proxy 
tr2 <- tr |>
  st_transform(25833)

mire <- read_stars(file) |>
  st_downsample(100) |>
  st_as_stars()
plot(mire, col="yellow")

tr_mire  <- stars::read_stars(file, proxy=T) |>
  st_crop(tr) |>
  st_as_stars()
#Not working

# What is the EDM is actually utm 33
tr_mire  <- stars::read_stars(file, proxy=T)
st_crs(tr_mire) <- 25833
tr_mire  <- tr_mire |>
  st_crop(tr2) |>
  st_as_stars()

@@

# This is not working as it puts the cropped raster into the sea
tmap_mode("view")
tm_shape(st_downsample(tr_mire , 10))+
  tm_raster(alpha=.8)+
  tm_shape(outline)+
  tm_borders()

@@

```{r}
#| eval: true

# calculate terrestrial area
terrestrial <- terrestrial |> 
  mutate(t_area_km = 
           round(units::drop_units(area_t * 1e-6))) |>
  as_tibble()

# calculate area of survey coverage maps
dk2 <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
  st_transform(st_crs(muni3)) |>
  st_intersection(muni3) |>
  st_cast("POLYGON") |>
  group_by(Municipality) |>
  summarise(geometry = st_union(geometry)) |>
  mutate(dk_area_km = geometry |> st_area(),
         dk_area_km = round(units::drop_units(dk_area_km * 1e-6)))

muni_tbl <- muni3 |>
  # calculate total area
  mutate(area_km = 
           round(
             units::drop_units(
               geom |> st_area() * 1e-6))) |>
  # make tibble for the coming join
  as_tibble() |>
  # paste inn terrestrial area
  left_join(terrestrial |> select(kommunenummer, t_area_km),
            keep=F) |>
  # area of survey
  left_join(dk2 |> select(Municipality, dk_area_km)) |>
  mutate(dk_percent = round((dk_area_km/t_area_km) * 100))

@@

| Municipality | Total terrestrial area (km2)                                                 | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area | Total mire area | Fraction mire area included in survey | Number of mire polygons in survey |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| Nordre Follo | `r terrestrial |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` |                                                     |                 |                                       |                                   |
| Trondheim    | `r terrestrial |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                     |                 |                                       |                                   |
| Målselv      | `r terrestrial |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                     |                 |                                       |                                   |

: Information for the three target municipalities in Norway
